---
title: "NYC Shooting Incident Data (2006-2022)"
author: "Omid Soufipour"
date: "2023-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Introduction
I conducted a data analysis of shooting reports in New York City for a course on the Master of Science in Data Science program at the **University of Colorado Boulder**. To complete the project, I utilized publicly available data from the [City of NewYork website](https://data.cityofnewyork.us/Public-Safety/
NYPD-Shooting-Incident-Data-Historic-/833y-fsy8). The purpose of this analysis was to gain insights into the frequency and location of shootings in the city.


## Libraries
I used only the **"tidyverse"** library for analyzing data and drawing plots. The library **"lubridate"** is also in the same package, but it needed to call seperately.


```{r library}
library(tidyverse)
library(lubridate)

```
  
## Data Analysis

  For the purpose of data analysis, I performed the following steps:

#### 1. Reading the data:
  I retrieved the data from the website mentioned earlier and saved it in a file named "nyc_shootings".

```{r read}
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nyc_shootings <- read_csv(url)

```

#### 2- Checking the data:
  I need to examine the data to determine the type of analysis and any necessary modifications.

```{r glimpse_data}
glimpse(nyc_shootings)

```



#### 3- Removing unwanted columns:

  After reviewing the columns, I decided to delete the following columns:


```{r remove_columns}
nyc_shootings <- nyc_shootings %>% 
  select(-LOCATION_DESC, -X_COORD_CD, -Y_COORD_CD, -Latitude,
         -Longitude, -Lon_Lat, -JURISDICTION_CODE, -PRECINCT)
```


#### 4- Renaming some columns:

It's a good idea to rename the column names to names that are more suitable for R programming.
  If you think it's the easy part, just read this quote from Phil Karlton:  
     ”*There are only two hard things in Computer Science: cache invalidation and naming things.*”


```{r rename_columns}
nyc_shootings <- nyc_shootings %>% 
  rename(date = OCCUR_DATE, time = OCCUR_TIME, borough = BORO,
         murder = STATISTICAL_MURDER_FLAG, 
         incident_key = INCIDENT_KEY, perpetrator_age = PERP_AGE_GROUP,
         perpetrator_sex = PERP_SEX, 
         perpetrator_race = PERP_RACE, victim_age = VIC_AGE_GROUP,
         victim_sex = VIC_SEX, victim_race = VIC_RACE)
```


#### 5- Other changes to the data frame:

  Now is a good opportunity to make additional modifications to the dataframe. First, I will update the "incident_key" column since the current data appears to be meaningless. My plan is to sort the data based on the date and assign a unique ID, starting from 1, to each incident.  
  Before proceeding with this task, I will convert the date format to a standard format.
  
  

```{r date_format}
nyc_shootings <- nyc_shootings %>% 
  mutate(date=mdy(date)) %>%
  arrange(time) %>% 
  arrange(date)
```

  Next, I added a id column that uses the date to create a meaningful identifier, and removed the incident_key column.


```{r id}
nyc_shootings <- nyc_shootings %>%
  mutate(id = dense_rank(incident_key))


unique_incident_key <- distinct(nyc_shootings,
  incident_key, .keep_all = TRUE)$incident_key

nyc_shootings$id <- match(nyc_shootings$incident_key, unique_incident_key)

nyc_shootings <- nyc_shootings %>% 
  select(-incident_key)
```

  I added two columns to the dataset - the first column indicating the weekday, and the second column indicating the time of day.

```{r time_day}
nyc_shootings <- nyc_shootings %>% 
  mutate(weekday = weekdays(date),
         time_of_day = case_when(hour(time) < 6 ~ "midnight",
                                 hour(time) < 12 ~ "morning",
                                 hour(time) < 18 ~ "afternoon",
                                 hour(time) < 24 ~ "evening"))
```



  I added the current population data (as of 2022) using the data from the [citypopulation website](http://www.citypopulation.de/en/usa/newyorkcity/)

```{r population}
nyc_shootings$population <- NA

nyc_shootings$population[nyc_shootings$borough == "BRONX"] <- 1424948
nyc_shootings$population[nyc_shootings$borough == "BROOKLYN"] <- 2641052
nyc_shootings$population[nyc_shootings$borough == "MANHATTAN"] <- 1576876
nyc_shootings$population[nyc_shootings$borough == "QUEENS"] <- 2331143
nyc_shootings$population[nyc_shootings$borough == "STATEN ISLAND"] <- 493494
```


  We have many empty cells, as well as a significant number of cells labeled 'U' or 'UNKNOWN.' Since there are more of these than we can ignore, I will replace all of them with the label 'unknown'.


```{r unknown}
nyc_shootings[, !(names(nyc_shootings) %in% c("date", "time"))] <- lapply(nyc_shootings[, !(names(nyc_shootings) %in% c("date", "time"))], function(x) ifelse(is.na(x), "unknown", x))
nyc_shootings[, !(names(nyc_shootings) %in% c("date", "time"))] <- replace(nyc_shootings[, !(names(nyc_shootings) %in% c("date", "time"))], nyc_shootings[, !(names(nyc_shootings) %in% c("date", "time"))] == "U", "unknown")
nyc_shootings[, !(names(nyc_shootings) %in% c("date", "time"))] <- replace(nyc_shootings[, !(names(nyc_shootings) %in% c("date", "time"))], nyc_shootings[, !(names(nyc_shootings) %in% c("date", "time"))] == "UNKNOWN", "unknown")


```



  I rearranged the order of the columns and removed the "time" column.

```{r order}
nyc_shootings <- nyc_shootings %>%
  select(id, everything()) %>% 
  select(-time)
```
  

## Plots

  I am beginning to create some plots. However, it would be beneficial to have a table displaying the number of shootings per day beforehand.


```{r table1}
shootings_per_day <- nyc_shootings %>%
  count(date) %>%
  arrange(desc(n))

shootings_per_day
```

  What a day July 5th, 2020 was! Unbelievable! There were 47 shootings in just one day!  
    Let's perform some more analysis. I am interested in determining which day of the week and time of day had the highest number of shootings. I would like to create a plot to visualize this information.
  
  

```{r plot1}
nyc_shootings %>%
  group_by(weekday, time_of_day) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = factor(weekday, levels = c("Monday", "Tuesday",
          "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
  y = count, fill = factor(time_of_day, levels = c("midnight","morning","afternoon","evening")))) +
  geom_col(position = "dodge") +
  labs(x = "Day of the Week", y = "Number of Shootings", fill = "Time of Day") +
  scale_fill_manual(values = c("morning" = "lightblue", "afternoon" = "orange", 
                               "evening" = "darkred", "midnight" = "black")) +
  theme_bw()+labs(title = "1- Number of shootings per weekday and time of day")



```  


chart of number of shootings per season:  

```{r plot2}
nyc_shootings %>%
  mutate(season = ifelse(month(date) %in% c(12, 1, 2), "Winter",
                         ifelse(month(date) %in% c(3, 4, 5), "Spring",
                                ifelse(month(date) %in% c(6, 7, 8), "Summer", "Fall")))) %>%
  group_by(season) %>%
  summarise(shootings = n()) %>%
  ggplot(aes(x = factor(season, levels=c("Spring", "Summer", "Fall", "Winter")), y = shootings, fill = season)) +
  geom_col() +
  labs(x = "Season", y = "Number of shootings", title = "2- Number of shootings per season") +
  theme_bw()
```



I want to compare the number of shootings across different boroughs in New York City.


```{r plot3}
NYPD_shooting_with_pop <- nyc_shootings %>%
  group_by(borough) %>%
  summarize(total_population = max(population),
            incidents_count = n(),
            percentage_of_shootings = incidents_count*1000/total_population)

NYPD_shooting_with_pop%>%
  ggplot(aes(x = borough, y = percentage_of_shootings)) +
  geom_bar(stat = "identity") +
  labs(x = "Location", y = "Number of shootings per 1000 people ",
       title = "3- Comparing NYC boroughs based on shootings")
```


  Now, I would like to create a table that displays the number of murders per day in descending order. As you may recall, the highest number of shootings in a single day occurred on July 5, 2020. However, according to the table below, this date ranks second in terms of the number of murders.

```{r killing}
nyc_shootings %>%
  filter(murder == TRUE) %>%
  group_by(date) %>%
  summarise(murders = n()) %>%
  arrange(desc(murders))
```


I want to create a plot that shows the number of shootings by gender.


```{r shooting_per_sex}
ggplot(nyc_shootings, aes(x = perpetrator_sex, fill = victim_sex)) +
  geom_bar(position = "dodge") +
  labs(title = "4- Shooting by Perpetrator Sex and Victim Sex",
       x = "Perpetrator Sex", y = "number of shootings") +
  theme_minimal()
```

I want to create a plot that shows the number of shootings by age group:


```{r shooting_per_age}
ggplot(nyc_shootings, aes(x = perpetrator_age, fill = victim_age)) +
  geom_bar(position = "dodge") +
  labs(title = "5- Shooting by Perpetrator Age and Victim Age",
       x = "Perpetrator Age", y = "number of shootings") +
  theme_minimal()
```


  If anyone is interested, they can use my previous code to conduct the same analysis on race. I excluded this analysis to prevent my project from becoming overly lengthy.



## Define a Model

  As the final step of my project, I aim to define a model that demonstrates the correlation between the number of shootings and murders per year.

```{r model}
shootings_by_year <- nyc_shootings %>%
  group_by(year = lubridate::year(date)) %>%
  summarize(shootings = n()) %>%
  ungroup()

murders_by_year <- nyc_shootings %>%
  filter(murder) %>%
  group_by(year = lubridate::year(date)) %>%
  summarize(murders = n()) %>%
  ungroup()

shootings_and_murders_by_year <- shootings_by_year %>%
  left_join(murders_by_year, by = "year") %>%
  mutate(murders = ifelse(is.na(murders), 0, murders))

lm_model <- lm(murders ~ shootings, data = shootings_and_murders_by_year)
summary(lm_model)

```

  and here is the plot of this simple model:

```{r model_plot}
shootings_and_murders_by_year %>%
  ggplot(aes(x = shootings, y = murders)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Shootings per Year", y = "Murders per Year",
       title = "Linear Model between Shootings and Murders in NYC") +
  theme_bw()

```



## Conclusion and potential biases

  Based on the project and the plots I have created, the following conclusions can be drawn:

1. Most shootings occurred from Friday evening until Sunday morning, indicating that they were not with prior intention and happened spontaneously.  
2. The plot showing the number of shootings per season confirms my earlier finding that most shootings were not premeditated. Summer had the highest number of shootings, which could be attributed to people going out to nightclubs and other recreational activities on weekends. This finding contradicts my initial assumption that crime groups were responsible for most shootings, as it appears that ordinary people are also involved in such incidents. However, it's important to acknowledge the possibility of biases in my analysis, and further research is needed to confirm these findings.  
3. The available age group and sex data do not allow for precise conclusions as there is a significant amount of unknown data. However, based on the available data, it can be said that the majority of both perpetrators and victims were male and in the 25-44 age group.
